

# EDA
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
```

```{r}
df<-read.csv("../App/2018-in-precinct-general.csv",header = TRUE)
colnames(df)
```
```{r}
df |> head(10)
```


```{r}
unique(df$party_detailed)
unique(df$party_simplified)
```

### lets look at each county and party votes at each county


```{r}
county_party_groupby <- df |> group_by(county_name, party_simplified)|> summarise(total_votes = sum(votes),.groups = "drop")

county_party_groupby
```

### lets visualize the results above





## lets resume again

```{r}
df<-read.csv("../App/2018-in-precinct-general.csv",header = TRUE)
df$party_simplified[is.na(df$party_simplified)] <- "OTHER"
counties <- unique(df$county_name) ## will sort later
parties <- unique(df$party_simplified)
## its better to store already group by-ed DF as well, since its quick filter 
county_party_groupby <- df |> group_by(county_name, party_simplified)|> summarise(total_votes = sum(votes),.groups = "drop")
```



```{r}
county_party_groupby
```

```{r}
colnames(df)
```



```{r}
unique(df$office)
```


```{r}
unique(df$magnitude)
```


```{r}
unique(df$mode)
```


## Questions I need to address

- *What are percentage of ABSENTEE and EARLY are present ?*
- Visualize the ABSENTEE, EARLY , PROVISIONAL or ELECTION DAY pie chart, overall , county-wise , party-wise
- (Side Note) There can be two types of visualizations, Aggregate of choosen options, and also comparsion among choosen options

- *What OFFICE has more major dominance by one party (competitive/one-sided) ? (I will make between DEMO and REPUB only)*
- Visualize Sorted bar/col plot with One-sided to competitive OFFICES, each of the bar colored based on DEMO and REPUB
- Allowing user to look at top N or choose what they are specifically looking for and also allowing user to sort based on One-sided-ness or competitive-ness
- Toggle between Votes Diff or Percentage Diff

- *Votes distribution across County Maybe Geographical Plot (is it possible?)*




```{r}
### Lets just care about few columns and use it for plotting of the above question
important_cols<- c("county_name","office","candidate","votes","mode","party_simplified")
df <- df[,important_cols]

### get the count and percentage of ABS, Ear etc
df |> count(mode) 
```

- This is interesting, I need to figure out a way to address these empty mode values, 
- **I can make them default which would be election day. Since its not election day , its likely that these would have been marked by Authorties** (Need to mention as SERIOUS bias)
- So lets convert them into ELECTION DAY mode

- OR Maybe its better to mark them UNKNOWN and exclude from plotting I suppose

```{r}
df <- df |> mutate(
  mode = ifelse(is.na(mode) | mode == "", "UNKNOWN",mode)
)
df |> count(mode) 
df |> count(party_simplified)
```

```{r}
## Suppose we choose county and party

SelectedCounty <- c("WHITLEY","MARION")
SelectedParty <- c("DEMOCRAT","REPUBLICAN")

##- *What are percentage of ABSENTEE and EARLY are present ?*

## for aggregated plot 
aggregate_mode_df <- df |> 
  filter(party_simplified %in% SelectedParty, county_name %in% SelectedCounty) |> 
      group_by(mode) |> summarise(total_votes = sum(votes),.groups = "drop") |> mutate(percent = total_votes*100/sum(total_votes))

## for comparison plot
comparison_mode_df <- df |> group_by(mode,party_simplified,county_name) |> 
  summarise(total_votes = sum(votes),.groups = "drop") |> filter(party_simplified %in% SelectedParty,  county_name %in% SelectedCounty) |> mutate(percent = total_votes*100/sum(total_votes))


```


```{r}
## Pie chart plotting 
aggregate_mode_df |> filter(mode != "UNKNOWN") |>
ggplot( aes(x = "", y = percent, fill = mode)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  labs(
    title = "Vote Share by Voting Mode",
    fill = "Voting Mode"
  ) + theme_void()
```

```{r}
comparison_mode_df |>
  filter(mode != "UNKNOWN") |>
  ggplot(aes(x = "", y = total_votes, fill = mode)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~ party_simplified) +
  theme_void() +
  labs(
    title = "Vote Share by Voting Mode by Party",
    fill = "Mode"
  )
```





- *What OFFICE has more major dominance by one party (competitive/one-sided) ? (I will make between DEMO and REPUB only)*
- Visualize Sorted bar/col plot with One-sided to competitive OFFICES, each of the bar colored based on DEMO and REPUB
- Allowing user to look at top N or choose what they are specifically looking for and also allowing user to sort based on One-sided-ness or competitive-ness
- Toggle between Votes Diff or Percentage Diff


```{r}
## groupby office , party and county
groupbyOffice_party_county <- df |> filter(party_simplified %in% c("DEMOCRAT","REPUBLICAN"), county_name %in% SelectedCounty) |>
      group_by(office,party_simplified,county_name) |> summarise(total_votes = sum(votes),.groups = "drop") 
        
```


```{r}
## aggregate county office diff(demo vs repub)

agg_office_diff_df <- df |> filter(party_simplified %in% c("DEMOCRAT","REPUBLICAN"), county_name %in% SelectedCounty) |>
      group_by(office,party_simplified) |> summarise(total_votes = sum(votes),.groups = "drop") |>  
    pivot_wider( names_from  = party_simplified,values_from = total_votes,values_fill = 0) |> ## pivoting only party simplified and total_votes 
  mutate(vote_diff = DEMOCRAT - REPUBLICAN)

```


```{r}
 ggplot(agg_office_diff_df, aes(x = office, y = vote_diff, fill = vote_diff > 0)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Democrat vs Republican Vote Difference by Office",
    y = "Vote Difference (DEM - REP)"
  ) +
  scale_fill_manual(values = c("red", "blue"), guide = "none")

```

### I will try to use Map , I am doing this first time, hopefully it works out

```{r}
library(sf)
library(tigris)
library(ggplot2)
library(dplyr)
```


```{r}
options(tigris_use_cache = TRUE)

indiana_counties <- counties(
  state = "IN",
  cb = TRUE,        
  year = 2022
)

```

```{r}
ggplot(indiana_counties) +
  geom_sf(fill = "white", color = "black") +
  labs(title = "Indiana Counties") +
  theme_void()

```

```{r}
#I will need fresh df since I used only important_cols for my analysis so far

new_df <- read.csv("../2018-in-precinct-general.csv",header = TRUE)

county_votes <- new_df |>
  group_by(county_fips) |>
  summarise(
    total_votes = sum(votes),
    .groups = "drop"
  )


county_votes$county_fips <- as.character(county_votes$county_fips)


map_df <- indiana_counties |>
  mutate(county_fips = GEOID) |>
  left_join(county_votes, by = "county_fips")

ggplot(map_df) +
  geom_sf(aes(fill = total_votes), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Total Votes by County (Indiana)",
    fill = "Votes"
  ) +
  theme_void()

```


```{r}
## Lets do DEMO vs REPUBLIC
county_demo_republican <- new_df |> filter(party_simplified %in% c("DEMOCRAT","REPUBLICAN")) |> group_by(county_name,county_fips,party_simplified) |> summarise(total_votes = sum(votes),.groups = "drop") |> pivot_wider(names_from = party_simplified, values_from = total_votes,values_fill = 0) |> mutate(vote_diff = DEMOCRAT - REPUBLICAN)

county_demo_republican
```

```{r}
county_demo_republican$county_fips <- as.character(county_demo_republican$county_fips)
map_df <- indiana_counties |>
  mutate(county_fips = GEOID) |>
  left_join(county_demo_republican, by = "county_fips")




ggplot(map_df) +
  geom_sf(aes(fill = vote_diff), color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#B22222",      # Republican red
    mid = "white",
    high = "#1E3A8A",     # Democrat blue
    midpoint = 0,
    na.value = "grey90",
    name = "Vote Margin\n(DEM − REP)"
  ) +
  theme_void() +
  labs(title = "Indiana County Vote Margin")

```

```{r}
#- *Candidate-wise Votes distribution across County Maybe Graphical Plot (is it possible?)*
df |> count(county_name) ## 92 counties!!!

```


### Lets test utils.R

```{r}
source("../App/utils.R")
df<-read.csv("../App/2018-in-precinct-general.csv",header = TRUE)
#County_total_votes_IN_MAP_df

x<- indiana_counties |> left_join(County_total_votes_IN_MAP(df)) |> ggplot() +
  geom_sf(aes(fill = total_votes), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Total Votes by County (Indiana)",
    fill = "Votes"
  ) +
  theme_void()
#plot_IN_MAP_TOTAL_VOTES(County_total_votes_IN_MAP_df, indiana_counties)
```




```{r}
library(plotly)
ggplotly(x)
```

```{r}
County_total_votes_IN_MAP(df)$county_fips
```

```{r}
indiana_counties$county_fips
```

```{r}
df |> count(county_name) |> arrange(desc(n))
```

wow,apparently, r has plotly, lets try

```{r}
df$party_simplified <- ifelse(is.na(df$party_simplified) | (df$party_simplified == ""),"OTHER",df$party_simplified)
Demo_republic_votes_diff_office_static <- df |> filter(party_simplified %in% c("DEMOCRAT","REPUBLICAN"))|>
  group_by(office,party_simplified,county_name) |> 
  summarise(total_votes = sum(votes),.groups = "drop") |>  
  pivot_wider( names_from  = party_simplified,values_from = total_votes,values_fill = 0) |> ## pivoting only party simplified and total_votes 
  mutate(vote_diff = DEMOCRAT - REPUBLICAN) 
Demo_republic_votes_diff_office_static
```

```{r}
SelectedCounty <- c("WHITLEY","MARION")
SelectedOffice <- c("US SENATE","US OFFICE","SECRETARY OF STATE","STATE TREASURER")
Demo_republic_votes_diff_office_static |>
  filter(county_name %in% SelectedCounty,office %in% SelectedOffice) |>
  group_by(office)|>
summarise(votes_diff = sum(vote_diff),.groups = "drop")
```



### Lets try to fetch from my server

```{r}
library(httr)
library(jsonlite)

url <- "http://100.110.163.111:8000/data/full"

res <- GET(url)
data <- fromJSON(content(res, "text"))
res
data
```

```{r}
apidf <- data
colnames(apidf)
```

```{r}
unique(apidf$`Office Category`)
```


## lets make ui better


```{r}
df<-read.csv("../App/2018-in-precinct-general.csv",header = TRUE)
df$party_simplified <- ifelse(is.na(df$party_simplified) | (df$party_simplified == ""),"OTHER",df$party_simplified)
counties <- unique(df$county_name) 
parties <- unique(df$party_simplified)
parties_map <- setNames(parties, parties)
offices <- unique(df$office)
```


```{r}
parties
map_parties = c()
for(i in parties){
  map_parties[i] = i
}
map_parties
```

```{r}
Total_Votes_per_Party <- df |> group_by(party_simplified) |> summarise(total_votes = sum(votes),.groups = "drop")
Total_Votes_per_Party[Total_Votes_per_Party$party_simplified == "DEMOCRAT",]$total_votes
```


